A Logical Solution of the Linearity Challenge

### 2025-08-01 paper/Process.lagda.tex
drafted deadlock freedom

The \AgdaInductiveConstructor{join} process has a single continuation which uses
the pair of continuation channels received from a channel of type $A \Par B$.
Note that $B$ is prepended in front of $A$, somewhat implying that the second
continuation is bound \emph{after} the first even though, from a technical
standpoint, both continuations are bound simultaneously. We could have prepended
the types $A$ and $B$ also in the opposite order, provided that the reduction
rule \RFork (formalized in \Cref{sec:semantics-agda}) is suitably adjusted.

>>>>

The \AgdaInductiveConstructor{join} process has a single continuation which uses
the pair of continuation channels received from a channel of type $A \Par B$.
Note that $B$ is prepended in front of $A$, somewhat implying that the second
continuation is bound \emph{after} the first even though, from a technical
standpoint, both continuations are bound simultaneously. We could have prepended
the types $A$ and $B$ also in the opposite order, provided that the reduction
rule \RFork (formalized in \Cref{sec:reduction-agda}) is suitably adjusted.

### 2025-08-01 paper/Reduction.lagda.tex
drafted deadlock freedom

  r-cong      : ∀{P R Q : Process Γ} → P ⊒ R → R ↝ Q → P ↝ Q
\end{code}
\end{AgdaAlign}
In the following it is also useful to work with the reflexive, transitive
closure of $\red$, which is defined below:

\begin{code}
data _⇒_ {Γ} : Process Γ → Process Γ → Set where
  refl  : ∀{P : Process Γ} → P ⇒ P
  tran  : ∀{P Q R : Process Γ} → P ↝ Q → Q ⇒ R → P ⇒ R
\end{code}

\begin{code}[hide]
Reducible : ∀{Γ} → Process Γ → Set
Reducible P = ∃[ Q ] P ↝ Q
\end{code}

>>>>

  r-cong      : ∀{P R Q : Process Γ} → P ⊒ R → R ↝ Q → P ↝ Q
\end{code}
\end{AgdaAlign}

### 2025-08-01 paper/calculus.tex
drafted deadlock freedom

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) peer endpoints are always used in complementary ways. This
latter property is dubbed \emph{well formedness} in the text of the challenge.
In this work we also consider \emph{deadlock freedom} which follows from the
fact that \Calculus is based on linear logic. As we will see, well formedness is
nothing but a particular form of deadlock freedom. We now formulate these
properties using the notation developed so far.

@@

\begin{theorem}[subject reduction]
    \label{thm:red}
    If $P \red Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
\end{theorem}

In order to formulate deadlock freedom, we need to introduce some terminology
for referring to processes that are unable to make any progress. Deadlock is not
simply the negation of reducibility: there are irreducible processes that are
not a deadlock because they would be able to reduce if put in a suitable
context. For example, $\Wait\x.P$ does not reduce, and yet it is not a deadlock
because it would be able to make progress when composed in parallel with
$\Close\x$. In general, even a process like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where
$x\ne y,z$ cannot be considered a deadlock, because the prefixes $\Wait\y$ and
$\Wait\z$ could be exposed using \SWait and possibly \SComm. Let us make all
this more precise.

We say that $P$ is a \emph{thread} if it is anything but a cut. In other words,
a thread is either a link or a process that starts with an input/output action
of some sort. Note that a thread may contain cuts, but these must all guarded
underneath the topmost action prefix.
%
We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
An observable process is a process that exposes an action on a free channel name
and therefore can interact through that channel, if put into some appropriate
context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
reducible process may perform a reduction step.
%
Now, a process is a \emph{deadlock} if it is neither observable nor reducible. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$.

\begin{theorem}[deadlock freedom]
    \label{thm:df}
    If $\wtp{P}\Context$, then $P$ is not a deadlock.
\end{theorem}

In order to formulate \emph{well formedness}~\cite{CarboneEtAl}, we define
\emph{reduction contexts} as partial processes with a single unguarded hole
$\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
\]

As usual, we write $\ReductionContext[P]$ for the process obtained by replacing
the hole in $\ReductionContext$ with $P$, noting that such replacement may
capture channel names that are bound in $\ReductionContext$ and occur free in
$P$.
%
Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
complementary ways, then their parallel composition $\Cut\x{P}{Q}$ is reducible
according to one of the principal cut reductions described in
\Cref{tab:semantics}. Therefore, an alternative (and more general) way of
formulating well formedness is this:
%
we say that $P$ is \emph{well formed} if $P \pcong
\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is not a
deadlock.

\begin{theorem}[type safety]
    \label{thm:type-safety}
    If $\wtp{P}\Context$, then $P$ is well formed.
\end{theorem}

>>>>

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) peer endpoints are always used in complementary ways. This
latter property is called \emph{well formedness} in the challenge. In this work
we also consider \emph{deadlock freedom}, which is more general than well
formedness and holds for \Calculus since its type system is based on linear
logic. We now formulate these properties using the notation developed so far.

@@

\begin{theorem}[subject reduction]
    \label{thm:red}
    If $P \red Q$ then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
\end{theorem}

In order to formulate deadlock freedom, we first need to introduce some
terminology for referring to processes that are unable to make any progress. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
process is unable to reduce (because a $\Close\x$ process is meant to interact
with a process of the form $\Wait\x.P$) and, more generally, it is unable to
interact regardless of the context in which it occurs because the $\Close\x$
sub-processes it contains are blocked on the channel $x$ bound by the cut.
%
In general, the property of being a deadlock is not simply the inability to
reduce: there are irreducible processes that are not a deadlock because they
would be able to reduce if put in a suitable context. For example, $\Wait\x.P$
does not reduce, and yet it is not a deadlock because it would be able to make
progress when composed in parallel with $\Close\x$. In general, even a process
like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a
deadlock, because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using
\SWait and possibly \SComm. Let us make all this more precise.

We say that $P$ is a \emph{thread} if it is anything but a cut. In other words,
a thread is either a link or a process that starts with an input/output action
of some sort. Note that a thread may contain cuts, but these cuts must all
guarded underneath the topmost action prefix.
%
We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
An observable process is a process that exposes an action on a free channel name
and therefore can interact through that channel, if put into some appropriate
context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
reducible process may perform a reduction step.
%
We say that a process is \emph{alive} if it is either observable or reducible
and that it is a \emph{deadlock} if it is not alive.

\begin{theorem}[deadlock freedom]
    \label{thm:df}
    If $\wtp{P}\Context$ then $P$ is alive.
\end{theorem}

In order to formulate \emph{well formedness}~\cite{CarboneEtAl24}, we define
\emph{reduction contexts} as partial processes with a single unguarded hole
$\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
\]

As usual, we write $\ReductionContext[P]$ for the process obtained by replacing
the hole in $\ReductionContext$ with $P$, noting that such replacement may
capture channel names that are bound in $\ReductionContext$ and occur free in
$P$.
%
Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
complementary ways, then their parallel composition $\Cut\x{P}{Q}$ is reducible
according to one of the principal cut reductions described in
\Cref{tab:semantics}. Therefore, an alternative (and more general) way of
formulating well formedness is this:
%
we say that $P$ is \emph{well formed} if $P \pcong
\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is alive.

\begin{theorem}[type safety]
    \label{thm:type-safety}
    If $\wtp{P}\Context$ then $P$ is well formed.
\end{theorem}
